Score-based tests of measurement invariance: Use in practice

### 2014-07-14 frontiers/IMPSpres-knitr.Rnw
change the empirical part name.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@355 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Revisions to IMPS slides
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@356 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\begin{frame}
  \frametitle{Hypotheses}
 Hypothesis of  measurement invariance and alternative:
  \begin{columns}
    \begin{column}{0.48\textwidth}
      \begin{block}{2 Groups}
        \begin{align}
          &H_0: \underset{i\in\left\{1:n\right\}}{\bm{\theta}_i} = 
          \bm{\theta}_0,\\
          &H_1:
           \begin{cases}
               \underset{i \in \left\{v_{1}\right\}}{\bm{\theta}_i} = 
               \bm{\theta}_{v_{1}} \\
               \underset{i \in \left\{v_{2}\right\}}{\bm{\theta}_i} = 
               \bm{\theta}_{v_{2}}, 
           \end{cases}
        \end{align}
       % \begin{center}
       %  where $\left\{v_{1}\right\}+\left\{v_{2}\right\}=\left\{1:n\right\}$
       % \end{center}
      \end{block}
  \end{column}
  \begin{column}{0.48\textwidth}
    \begin{block}{G Groups}
         \begin{align}
          &H_0: \underset{i\in\left\{1:n\right\}}{\bm{\theta}_i} = 
                \bm{\theta}_0,\\
          &H_1: \underset{i \in \left\{v_{g}\right\}}{\bm{\theta}_i} = 
                \bm{\theta}_{v_{g}}, 
         \end{align}
      % \begin{center}
      %  where $\sum_{g=1}^{G}\left\{v_{g}\right\}=\left\{1:n\right\}$
      % \end{center}   
       $V_{g}$ can be categorical, \textit{ordinal} and continous.
    \end{block}
  \end{column}
\end{columns}

@@

\begin{frame}
  \frametitle{Cumulative Sum of Scores}
      \begin{itemize}
        \item Order individuals by an auxiliary variable.
        \item Decorrelate scores, using an estimate of the information
          matrix.    
      \end{itemize}
\begin{equation} 
    \label{eq:cumscore}
  {\bm B}(t; \hat {\bm \theta}) ~=~ \hat {\bm I}^{-1/2} n^{-1/2}
    \sum_{i = 1}^{\lfloor n \cdot t \rfloor} {\bm s}(\hat {\bm \theta}; x_{(i)})
  \qquad (0 \le t \le 1)
\end{equation}

\vspace{0.2in}
where $\hat{\bm I}$ is an estimate of the information matrix, $\lfloor
nt \rfloor$ is the integer part of $nt$, $t$ is the cumulative proportion 
of $n$, $t \in (1/n, n)$.
  \end{frame}

@@

     \begin{column}{0.33\textwidth}
          \begin{block}{\scriptsize{Empirical Cumulative Score}}
          \resizebox{3.3cm}{!}{
           $
            \begin{blockarray}{lccccc}
              \text{obs}& \theta_{1} & \theta_{2} & \ldots 
                     &\textcolor{red}{\theta_{k}} & \text{age} \\
               \begin{block}{l(cccc)c}
                 x_{1} &  1.32  & -4.32 & \ldots & -2.34 & 10.5 \\
                 x_{2} &  4.69 & 4.00 & \ldots & 61.11 & 11.0 \\
                 x_{3} &  8.81 & 7.48 & \ldots & 58.00 & 11.3 \\
                 x_{..} & \vdots & \vdots & \vdots & \vdots & 17.1 \\
                 x_{n-1} &  7.89 & 3.32 & \ldots & 4.91 & 18.6 \\
                 x_{n}& 0     &   0     & \ldots & 0 & 18.9 \\
               \end{block}
            \end{blockarray} 
           $
         }  
        \end{block}
      \end{column}
   \end{columns}
\end{frame}

@@

\begin{frame}
   \frametitle{Test Statistics}
\begin{itemize}
  \item Categorical $V$: Maximum Lagrange multiplier test (asymptotically equvialent to LRT).
\item Ordinal $V$: Weighted double maximum test ($\mathit{WDM}_o$);\\
Maximum Lagrange multiplier test ($\max\mathit{LM}_o$).
\item Continuous $V$: Double maximumum test ($\mathit{DM}$)\\
Cram{\'e}r-von Mises type statistic ($\mathit{CvM}$)\\
maximum Lagrange multiplier ($\max\mathit{LM}$)\\
\end{itemize}
\end{frame}

@@

\section{Simulation}
\begin{frame}
  \frametitle{Study Motivation}
  \begin{block}{Practical Considerations}
    \begin{itemize}
      \item Model misspecification is inevitable in practice.  
      \item Measurement invariance is often studied with respect to an
        \textbf{ordinal} variable.
      \item Availability/Convenience of software implementation.
    \end{itemize}
  \end{block}
\end{frame}

@@

\begin{frame}
  \frametitle{Conclusion based on Simulation Studies}
 \begin{itemize}
   \item
      The ordinal test statistics have better power 
      for detecting measurement 
      invariance violations that are monotonic 
      with $V$.
   \item This feature is robust as long as the violating 
      parameter is included in the estimated model.
   \item Use of these tests is convenient, requiring estimation of
     only a reduced model.
   \end{itemize}
\end{frame}

@@

\section{Application}
\begin{frame}
  \frametitle{Software Illustration}
The tests can be carried out in R. Specifically, 
\begin{itemize}
\item Model estimation can be carried out in \emph{lavaan} package. 
\item The test statistic and plot is obtained from \emph{strucchange} package.
\end{itemize}
\end{frame}

@@

\begin{frame}
\begin{figure}
\tiny{\caption{Fluctuation processes for $\max\mathit{LM}_o$ statistic}}
\label{fig:ordres}
<<sctest-ord-graphics,  out.width="250px", out.height ="180px", echo=FALSE, results='hide'>>=
sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
  functional = maxLMo, plot = TRUE,  
               ylim = c(0, 14), xlab = "Age group", 
               ylab = "LM statistics", main = expression(maxLM[o]))
@
\end{figure}
\end{frame}

@@

\begin{frame}
 \begin{itemize}
 \item Thanks! 
 \item Questions?
 \end{itemize}
\vspace{0.3in}
 Wang, T., Merkle, E. C., Zeileis, A. (2014). Score-based tests of measurement 
 invariance: Use in practice. \emph{Frontiers in Psychology}, {\em 5(438)}, 
 1--11. 
\end{frame}

>>>>

\begin{frame}
  \frametitle{Hypotheses}
 Hypothesis of  measurement invariance and alternative:
  \begin{columns}
    \begin{column}{0.48\textwidth}
      \begin{block}{2 Groups}
        \begin{align}
          &H_0: \underset{i\in\left\{1:n\right\}}{\bm{\theta}_i} = 
          \bm{\theta}_0,\\
          &H_1:
           \begin{cases}
               \underset{i \in \left\{v_{1}\right\}}{\bm{\theta}_i} = 
               \bm{\theta}_{v_{1}} \\
               \underset{i \in \left\{v_{2}\right\}}{\bm{\theta}_i} = 
               \bm{\theta}_{v_{2}}, 
           \end{cases}
        \end{align}
       % \begin{center}
       %  where $\left\{v_{1}\right\}+\left\{v_{2}\right\}=\left\{1:n\right\}$
       % \end{center}
      \end{block}
  \end{column}
  \begin{column}{0.48\textwidth}
    \begin{block}{G Groups}
         \begin{align}
          &H_0: \underset{i\in\left\{1:n\right\}}{\bm{\theta}_i} = 
                \bm{\theta}_0,\\
          &H_1: \underset{i \in \left\{v_{g}\right\}}{\bm{\theta}_i} = 
                \bm{\theta}_{v_{g}}, 
         \end{align}
      % \begin{center}
      %  where $\sum_{g=1}^{G}\left\{v_{g}\right\}=\left\{1:n\right\}$
      % \end{center}   
       $V_{g}$ can be categorical, \textit{ordinal}, or continuous.
    \end{block}
  \end{column}
\end{columns}

@@

\begin{frame}
  \frametitle{Empirical Cumulative Score Process}
      \begin{itemize}
        \item Order individuals by an auxiliary variable.
        \item Decorrelate scores, using an estimate of the information
          matrix.    
\begin{equation} 
    \label{eq:cumscore}
  {\bm B}(t; \hat {\bm \theta}) ~=~ \hat {\bm I}^{-1/2} n^{-1/2}
    \sum_{i = 1}^{\lfloor n \cdot t \rfloor} {\bm s}(\hat {\bm \theta}; x_{(i)})
  \qquad (0 \le t \le 1)
\end{equation}

\vspace{0.2in}
where $\hat{\bm I}$ is an estimate of the information matrix, $\lfloor
nt \rfloor$ is the integer part of $nt$, $t$ is the cumulative proportion 
of $n$, $t \in (1/n, n)$.
  \end{itemize}
  \end{frame}

@@

     \begin{column}{0.33\textwidth}
          \begin{block}{Cumulative Score}
          \resizebox{3.3cm}{!}{
           $
            \begin{blockarray}{lccccc}
              \text{obs}& \theta_{1} & \theta_{2} & \ldots 
                     &\textcolor{red}{\theta_{k}} & \text{age} \\
               \begin{block}{l(cccc)c}
                 x_{1} &  1.32  & -4.32 & \ldots & -2.34 & 10.5 \\
                 x_{2} &  4.69 & 4.00 & \ldots & 61.11 & 11.0 \\
                 x_{3} &  8.81 & 7.48 & \ldots & 58.00 & 11.3 \\
                 x_{..} & \vdots & \vdots & \vdots & \vdots & 17.1 \\
                 x_{n-1} &  7.89 & 3.32 & \ldots & 4.91 & 18.6 \\
                 x_{n}& 0     &   0     & \ldots & 0 & 18.9 \\
               \end{block}
            \end{blockarray} 
           $
         }  
        \end{block}
      \end{column}
   \end{columns}
\end{frame}

@@

\begin{frame}
   \frametitle{Test Statistics}
\begin{itemize}
    \itemsep8pt
  \item Categorical $V$: Maximum Lagrange multiplier test (asymptotically equvialent to LRT).
\item Ordinal $V$: Weighted double maximum test ($\mathit{WDM}_o$);\\
Maximum Lagrange multiplier test ($\max\mathit{LM}_o$).
\item Continuous $V$: Double maximum test ($\mathit{DM}$)\\
Cram{\'e}r-von Mises type statistic ($\mathit{CvM}$)\\
maximum Lagrange multiplier ($\max\mathit{LM}$)\\
\end{itemize}
\end{frame}

@@

\section{Simulation}
\begin{frame}
  \frametitle{Study Motivation}
  \begin{block}{Practical Considerations}
    \begin{itemize}
      \item Model misspecification is inevitable in practice.  
      \item Measurement invariance is often studied with respect to an
        \textbf{ordinal} variable.
      \item Availability/convenience of software implementation.
    \end{itemize}
  \end{block}
\end{frame}

@@

\begin{frame}
  \frametitle{Conclusion based on Simulation Studies}
 \begin{itemize}
     \itemsep8pt
   \item
      The ordinal test statistics have better power 
      for detecting measurement 
      invariance violations that are monotonic 
      with $V$.
   \item This feature is robust as long as the violating 
      parameter is included in the estimated model.
   \item Use of these tests is convenient, requiring estimation of
     only a reduced model.
   \end{itemize}
\end{frame}

@@

\section{Application}
\begin{frame}
  \frametitle{Software Illustration}
The tests can be carried out in R. Specifically, 
\begin{itemize}
\item Model estimation via package \emph{lavaan}. 
\item The test statistic and plot via package \emph{strucchange}.
\end{itemize}
\end{frame}

@@

\begin{frame}
\begin{figure}
\tiny{\caption{Fluctuation processes for $\max\mathit{LM}_o$ statistic}}
\label{fig:ordres}
<<sctest-ord-graphics, out.width="2.5in", out.height ="2.5in", echo=FALSE, results='hide', fig.height=5, fig.width=5>>=
sctest(restr, order.by = yg$agegroup, parm = 1:4, vcov = "info",
  functional = maxLMo, plot = TRUE,  
               ylim = c(0, 14), xlab = "Age group", 
               ylab = "LM statistics", main = expression(maxLM[o]))
@
\end{figure}
\end{frame}

@@

\begin{frame}
 \begin{itemize}
 \item Thanks! 
 \end{itemize}
\vspace{0.3in}
 Wang, T., Merkle, E. C., Zeileis, A. (2014). Score-based tests of measurement 
 invariance: Use in practice. \emph{Frontiers in Psychology}, {\em 5(438)}, 
 1--11. 
\end{frame}
